Interactive 3D Graphics of PITCHf/x Data
=======================================

<!--- view local page:
browseURL(paste0('http://127.0.0.1:', tools:::httpdPort, '/library/pitchRx/doc/index.html'))
-->

Good thing I follow everything my current T.A. [Yihui Xie](http://yihui.name) does when he isn't grading my bayes homework. Otherwise, I wouldn't have known about the wonderful [rgl](http://cran.r-project.org/web/packages/rgl/index.html) package - and it's [seamless html integration via knitr](https://twitter.com/xieyihui/status/302513617906044930).

I am absolutely giddy over the features rgl will add to my recently released R package [pitchRx](http://cran.r-project.org/web/packages/pitchRx/). Note that this example depends on developmental version of [pitchRx on github](https://github.com/cpsievert/), so if you're on windows, install [RTools](http://cran.rstudio.com/bin/windows/Rtools/) before installing [devtools](http://cran.r-project.org/web/packages/devtools/index.html).


```{r load, message=FALSE}
library(devtools)
library(knitr) #I'm using version 1.1
install_github('pitchRx', 'cpsievert')
library(pitchRx)
```

```{r setup, echo=FALSE, message=FALSE}
library(rgl)
knit_hooks$set(webgl = hook_webgl)
```

After pitchRx is loaded, I can bring in the sample dataset, then restrict to pitches thrown by Mariano Rivera's.  Next, a simple call to `rglFX` automatically generates an interactive 3D scatterplot (hold and drag your mouse on the scatterplot below - you can also zoom in and out). **rgl** currently has no support for including legends, so `rglFX` also prints the appropriate hex color codes in your R console.

```{r demo1, eval=FALSE}
interactiveFX(Rivera, spheres=FALSE, alpha=.5)
```

```{r webgl1, webgl=TRUE, echo=FALSE}
#ugly workaround since knitr won't recognize interactiveFX
Rivera <- subset(pitches, pitcher_name=="Mariano Rivera")
snaps <- getSnapshots(Rivera, interval = 0.01)
types <- as.character(Rivera[,"pitch_type"])
ncolors <- length(unique(types))
pal <- c(rgb(1, 0, 0), rgb(0, 0, 1))
full.pal <- factor(types)
levels(full.pal) <- pal
plot3d(x=as.vector(snaps[,,1]), y=as.vector(snaps[,,2]), z=as.vector(snaps[,,3]),
           xlab="Horizontal Axis", ylab="Distance from Home Plate", zlab="Height From Ground",
           col=as.character(full.pal), alpha=.5)
```

For a proper plotting perspective...

```{r demo2, eval=FALSE}
interactiveFX(Rivera, interval=.05, alpha=.75)
```

```{r webgl2, webgl=TRUE, echo=FALSE}
snaps <- getSnapshots(Rivera, interval = 0.05)
types <- as.character(Rivera[,"pitch_type"])
ncolors <- length(unique(types))
pal <- c(rgb(1, 0, 0), rgb(0, 0, 1))
full.pal <- factor(types)
levels(full.pal) <- pal
open3d()
spheres3d(x=as.vector(snaps[,,1]), y=as.vector(snaps[,,2]), z=as.vector(snaps[,,3]),
           col=as.character(full.pal), radius=.12, alpha=.5)
axes3d(c('x', 'y', 'z')) 
title3d(xlab='Horizontal Axis', ylab='Distance from Home Plate', zlab='Height From Ground')
```

I can also control how I want to color...

```{r demo3, eval=FALSE}
interactiveFX(Rivera, interval=.05, color="type", alpha=.75)
```

```{r webgl3, webgl=TRUE, echo=FALSE}
types <- as.character(Rivera[,"type"])
ncolors <- length(unique(types))
pal <- c(rgb(1, 0, 0), rgb(0, 1, 0), rgb(0, 0, 1))
full.pal <- factor(types)
levels(full.pal) <- pal
open3d()
spheres3d(x=as.vector(snaps[,,1]), y=as.vector(snaps[,,2]), z=as.vector(snaps[,,3]),
           col=as.character(full.pal), radius=.12, alpha=.75)
axes3d(c('x', 'y', 'z')) 
title3d(xlab='Horizontal Axis', ylab='Distance from Home Plate', zlab='Height From Ground')
```

This is a really great way to visualize PITCHf/x. With this tool, we can easily examine any sequence of pitches - from any angle. This particular graphic portrays the deception of Rivera's cut fastball compared to his four-seam.


#### Authored by <a href="http://cpsievert.github.com/home.html">Carson Sievert</a> via <a href="http://www.rstudio.com/ide/docs/authoring/using_markdown">R Markdown, knitr & RStudio</a>. Published by <a href="http://pages.github.com/">GitHub Pages</a>.